### 2018-08-16 paper/finitecomplete-failures.tex
small changes to Section 3, but a couple of notes that need attention

\begin{lemma}
\label{lemma:extendV}
Let $P$ be a CSP process with normalised transition graph $G(P)$, 
such that all states in $N$ are reachable.
Let $V\subseteq\Sigma^*$ be a finite prefix-closed set of
sequences of events. Suppose that  $G(P)$ reaches $k < |N|$ nodes under $V$, that is,
$|t(\ii n,V)| = k$. Let $V.\Sigma$ denote the set of all sequences from $V$, 
extended by any event of $\Sigma$.
Then $G(P)$ reaches at least $(k+1)$ nodes under $V\cup V.\Sigma$.
\end{lemma}
\begin{proof}
Suppose that $n'\in (N - t(\ii n,V))$.  Since all nodes in $N$ are reachable, there exists
a trace $s$ such that $G(P)/s = n'$. Decompose $s = s_1.e.s_2$ with $s_i\in\Sigma^*, e\in\Sigma$, such that $G(P)/s_1 \in t(\ii n,V)$ and $G(P)/s_1.e \not\in t(\ii n,V)$. Such a decomposition always exists, because $V$ is prefix-closed and therefore contains the empty trace $\varepsilon$. Note, however, that it is not necessarily the case that $s_1\in V$.

Since $G(P)$ reaches $G(P)/s_1$ under $V$, there exists a trace $u\in V$ such that
$\ol n = G(P)/u = G(P)/s_1$. Since $s = s_1.e.s_2$ is a trace of $P$ and 
$\ol n = G(P)/s_1$,
$(\ol n,e)$ is in the domain of $t$. Therefore,  
$n = G(P)/u.e = G(P)/s_1.e \not\in t(\ii n,V)$ is a well-defined node
of $N$ which is not contained in $t(\ii n,V)$. Since $u.e\in V\cup V.\Sigma$,
$G(P)$ reaches at least the additional node $n$ under $V\cup V.\Sigma$. This completes the proof.
\xbox
\end{proof}

@@

For a given reference process $P$ and for each integer $p\ge 0$, 
we define a CSP test process as follows. 
\begin{eqnarray}
U_F(p) & = & U_F(p,\varepsilon)
\\
U_F(p,s) & = & \big( e:(\Sigma - [P/s]^0) \then \efail\then \Stop \big)
\label{eq:ufa}
\\ & & \extchoice \nonumber
\\ & & ([P/s]^0 = \varnothing)    \&   \big( \epass \then \Stop \big)
\label{eq:ufb}
\\ & & \extchoice \nonumber
\\ & & (\#s < p) \& \big( e:[P/s]^0 \then U_F(p,s.e) \big)
\label{eq:ufc}
\\ & & \extchoice \nonumber
\\ & & (\#s = p) \& \big( \sqcap_{H\in\text{minHit}(P/s)} ( e:H \then \epass \then\Stop   )  \big)
\label{eq:ufd}
\end{eqnarray}

A test is performed by running $U_F(p)$ concurrently with any SUT process $Q$
that operates on the same alphabet as $P$, synchronising over
alphabet $\Sigma$. Therefore, a test execution is any trace of the concurrent process
\[
Q\parallel[\Sigma] U_F(p).
\]
It is assumed that the events $\efail$ and $\epass$ denoting FAIL and PASS of the test execution, respectively, are events outside $\Sigma$.
Since we assume that $Q$ is free of livelocks, it is guaranteed that each test execution terminates after some $s\in\trc(P)$ with length $(p+1)$ at the latest. 
The test is \emph{passed} by the SUT (we write $Q\ \pass\ U_F(p)$) if and only if {\it every} execution of $Q\parallel[\Sigma] U_F(p)$ terminates with PASS event $\epass$. This can also be  expressed by means of a failures refinement relation:
\[
Q\ \pass\ U_F(p) \equiv (\epass\then\Stop) \lessdet_F (Q\parallel[\Sigma] U_F(p)) \hide \Sigma  
\]
This type of pass relation is often called \emph{must test}, because every test execution must end with the $\epass$-event \cite{Hennessy:1988:ATP:50497}. Note that it
is necessary to use the failures refinement relation in this condition, and not the trace refinement relation: process $(Q\parallel[\Sigma] U_F(p)) \hide \Sigma$ may have  the same visible traces $\varepsilon, \epass$ as the ``Test Passed Process'' $(\epass\then\Stop)$. However,
the former may nondeterministically refuse $\epass$, due to a deadlock occurring when
a faulty SUT process executes concurrently with $U_F(p,s)$ executing branch (\ref{eq:ufd}),
because $\#s = p$. This is explained further in the next paragraphs.

Intuitively speaking, $U_F(p)$ is able to perform any trace $s$ of $P$, 
up to a length of $p$. 
If, after having already run through $s\in\trc(P)$ with $\#s < p$,
an event is accepted by the SUT which is outside the initials of $P/s$, the test 
immediately terminates with FAIL-event $\efail$. This is handled by the first operand of the
external choice in sub-process $U_F(p,s)$.

If $P/s$ is the STOP-process, this is revealed by its initials being empty. In this case, the test may terminate 
successfully (second operand of the external choice in $U_F(p,s)$). Note that at the same time, any (now illegal) event of the alphabet is also accepted by the test.
Therefore,
if the SUT would still accept an event in a state where $P/s$ is supposed to have stopped, there exists a test execution which terminates with FAIL by choosing the first
operand of the external choice.

If the length of $s$ is still less than $p$, the test accepts any event from the initials
$[P/s]^0$ and continues recursively as sub-process $U_F(p,s.e)$ (third operand of the external choice). A test of this type is called \emph{adaptive}, because it accepts any legal behaviour of the SUT and adapts its consecutive behaviour to the event selected by the SUT.

@@

\begin{theorem}\label{th:failurestest}
Let $P$ be a divergence-free CSP process over alphabet $\Sigma$ 
whose normalised transition graph $G(P)$ has $p$ states. Define fault domain ${\cal D}$ as
the set of all divergence-free CSP processes over alphabet $\Sigma$, whose transition graph
has at most $q$ states with $q \ge p$. 
Then the test suite 
\[
\TS_F = \{ U_F(k)~|~0 \le k < pq  \}
\]
is complete with respect to ${\cal F} = (P,\lessdet_F,{\cal D})$.
\end{theorem}
\begin{proof}
To prove soundness of $\TS_F$, suppose that $Q\in{\cal D} \wedge P\lessdet_F Q$. Then
$\trc(Q)\subseteq \trc(P)$, and Lemma~\ref{lemma:hseta} implies that 
for all traces $s$ of $Q$, every $H\in\text{minHit}(P/s)$ is a 
    hitting set for $\minaccs(Q/s)$. As a consequence, when running in parallel
with $Q$, any adaptive test $U_F(p)$ will always enter the branches 
(\ref{eq:ufb}), (\ref{eq:ufc}),  or (\ref{eq:ufd}) of the external choice construction
for $U_F(p,s)$. Branch (\ref{eq:ufb}) always leads to a PASS verdict, branch (\ref{eq:ufc})
always to test continuation without a verdict. For the last branch, we note that 
any selected minimal hitting set $H\in\text{minHit}(P/s)$ has a non-empty intersection with
each of the minimal acceptances of $Q/s$. As a consequence, $Q/s$ will never block when offered events from $H$, and the test terminates with PASS event $\epass$. Note that this argument requires that $Q$ is free of livelocks, because otherwise the PASS-events might not become visible, due to unbounded sequences of hidden events performed by $Q$. Note further, that this proof did not refer in any way to the size of $Q$'s normalised transition graph, so the test suite is sound for {\it all} non-divergent CSP process refining $P$.

To prove exhaustiveness, consider a process $Q\in{\cal D}$ with $P\not\lessdet_F Q$. This non-conformance can be caused in two possible ways.
\begin{description}
\item[Case~1] $\trc(Q)\not\subseteq \trc(P)$
\item[Case~2] There exists a joint trace $s\in\trc(Q)\cap\trc(P)$ and a minimal acceptance $A_Q$
of $\minaccs(Q/s)$, such that 
(see Lemma~\ref{lemma:tgtrcref}, (\ref{eq:failrefb})).
\begin{equation}
\label{eq:accsnotcontained}
\forall A_P\in\minaccs(P/s): A_P\not\subseteq A_Q,
\end{equation} 
\end{description}
It has to be shown for each of the two possibilities that at least one test execution of some $(Q\parallel[\Sigma] U_F(k))$ with $k < pq$ ends with the FAIL event $\efail$ or without giving any verdict. The latter case is also interpreted   as FAIL, since then  
$(\epass\then\Stop)$ is no longer refined by the test execution in the failures model.

For the first case, consider a  trace $s.e\in\trc(Q)$ such that $s\in\trc(P)$, but 
$s.e\not\in\trc(P)$. Then $s$ is also a trace of the product graph $G = G(P)\times G(Q)$ defined in Section~\ref{sec:ntg}. From the construction of $G$ described there, we know that $G$ has at most $pq$ reachable states, because $G(P)$ has $p$ states, and $G(Q)$ has at most $q$ states. Suppose that $G/s = (n_P,n_Q)$. Applying Lemma~\ref{lemma:extendV} implies that this state can be reached by a trace $u\in\trc(G)$ of length $\#u < pq$. Now the construction of the transition function of $G$ implies that $u$ is also a trace of $P$ and $Q$. Since test $U_F(pq-1)$ accepts all traces of $P$ up to length $pq-1$, $u$ is also a trace of this test, and, by construction, $U_F(pq-1)/u = U_F(pq-1,u)$. Since $s.e\not\in\trc(P)$,
$e$ is an element of $\Sigma-[P/u]^0$. Therefore, in at least one execution, $U_F(pq-1,u)$ executes its first branch (\ref{eq:ufa}) with this event $e$, so that the test fails with event $\efail$. Again, the assumption of non-divergence of Q is needed for this conclusion.

For the Case~2, we note that trace $s$ is again a trace of the product graph $G$. Therefore, $G/s$ can again be reached by a trace $u\in\trc(Q)\cap\trc(P)$ of maximal length
$\#u < pq$. Consider test $U_F(\# u)$ which satisfies $U_F(\# u)/u = U_F(\#u,u)$, because
it always performs branch (\ref{eq:ufc}) until the trace $u$ has been completely processed. $U_F(\#u,u)$
  may execute branches (\ref{eq:ufa}) or (\ref{eq:ufd}) only: 
Assumption (\ref{eq:accsnotcontained}) in Case~2 implies that $P/s$ has at least one non-empty minimal acceptance, so the guard condition $([P/s]^0 = \varnothing)$ of branch
 (\ref{eq:ufb}) evaluates to $\isf$ for $U_F(\#u,u)$. Moreover, 
 the guard condition $(\#s < p)$
for branch (\ref{eq:ufc}) evaluates to $\isf$ for $U_F(\#u,u)$, too. 
If branch (\ref{eq:ufa}) is executed, the test always fails. If branch (\ref{eq:ufd})
is executed, the test fails for the execution where a minimal hitting set 
$H\in\text{minHit}(P/u)$ is chosen by $U_F(\#u,u)$ which has an empty intersection
with the minimal acceptance $A_Q$ from condition (\ref{eq:accsnotcontained}). The existence of such an $H$ is guaranteed because of Lemma~\ref{lemma:hseta}. As a consequence, there exists 
a test execution   where $Q/u$ selects acceptance $A_Q$ and  $U_F(\#u,u)$ 
selects $H$. This execution deadlocks in 
process state $(Q\parallel[\Sigma]U_F(\# u))/u$, so it cannot 
produce the pass event $\epass$; this  means that the test fails. This concludes the proof.
\xbox
\end{proof}

>>>>

\begin{lemma}
\label{lemma:extendV} Let $P$ be a CSP process with normalised transition
graph $G(P)$, such that all states in $N$ are reachable. \fixme{alcc: can a
normalised graph for a process have states that are unreachable?} Let
$V\subseteq\Sigma^*$ be a finite prefix-closed set of sequences of events.
Suppose that  $G(P)$ reaches $k < |N|$ nodes under $V$, that is, $|t(\ii
n,V)| = k$. Let $V.\Sigma$ denote the set of all sequences from $V$, extended
by any event of $\Sigma$. Then $G(P)$ reaches at least $(k+1)$ nodes under
$V\cup V.\Sigma$.
\end{lemma}
\begin{proof}
Suppose that $n'\in (N - t(\ii n,V))$.  Since all nodes in $N$ are reachable,
there exists a trace $s$ such that $G(P)/s = n'$. Decompose $s = s_1.e.s_2$
with $s_i\in\Sigma^*, e\in\Sigma$, such that $G(P)/s_1 \in t(\ii n,V)$ and
$G(P)/s_1.e \not\in t(\ii n,V)$. Such a decomposition always exists, because
$V$ is prefix-closed and therefore contains the empty trace $\varepsilon$.
Note, however, that it is not necessarily the case that $s_1\in V$.

Since $G(P)$ reaches $G(P)/s_1$ under $V$, there exists a trace $u\in V$ such
that $G(P)/u = G(P)/s_1 = \ol n$. Since $s = s_1.e.s_2$ is a trace of $P$ and
$G(P)/s_1 = \ol n$, then $(\ol n,e)$ is in the domain of $t$. So, $ G(P)/u.e
= G(P)/s_1.e = n$ is a well-defined node of $N$ not contained in $t(\ii
n,V)$. Since $u.e\in V\cup V.\Sigma$, $G(P)$ reaches at least the additional
node $n$ under $V\cup V.\Sigma$. This completes the proof. \xbox
\end{proof}
\fixme{alcc: Explain the structure of the section? Can the above result be in
a section, rather than the introduction of the section?}

@@

For a given reference process $P$ and for each integer $p\ge 0$, we define a
CSP test process for failures refinement as shown below.
%
\begin{eqnarray}
U_F(p) & = & U_F(p,\varepsilon)
\\
U_F(p,s) & = & \big( \Extchoice e:(\Sigma - [P/s]^0) @ e \then \efail\then \Stop \big)
\label{eq:ufa}
\\ & & \extchoice \nonumber
\\ & & ([P/s]^0 = \varnothing)    \&   \big( \epass \then \Stop \big)
\label{eq:ufb}
\\ & & \extchoice \nonumber
\\ & & (\#s < p) \& \big( \Extchoice e:[P/s]^0 @ e \then U_F(p,s.e) \big)
\label{eq:ufc}
\\ & & \extchoice \nonumber
\\ & & (\#s = p) \& \big( \sqcap_{H\in\text{minHit}(P/s)} ( \Extchoice e:H @ e \then \epass \then\Stop   )  \big)
\label{eq:ufd}
\end{eqnarray}
%
A test is performed by running $U_F(p)$ concurrently with any SUT process $Q$
\fixme{alcc: add an example?} that operates on the same alphabet as $P$,
synchronising over alphabet $\Sigma$. Therefore, a test execution is any
trace of the concurrent process
\[
Q\parallel[\Sigma] U_F(p).
\]
It is assumed that the events $\efail$ and $\epass$, denoting FAIL and PASS
of the test execution, are events outside $\Sigma$. Since we assume that $Q$
is free of livelocks, it is guaranteed that each test execution terminates
after some $s\in\trc(P)$ with length $(p+1)$ at the latest. The test is
\emph{passed} by the SUT (written $Q\ \pass\ U_F(p)$) if, and only if, {\it
every} execution of $Q\parallel[\Sigma] U_F(p)$ terminates with PASS event
$\epass$. This can also be  expressed by means of a failures refinement.
\[
Q\ \pass\ U_F(p) \equiv (\epass\then\Stop) \lessdet_F (Q\parallel[\Sigma] U_F(p)) \hide \Sigma
\]
This type of pass relation is often called \emph{must test}, because every
test execution must end with the $\epass$
event~\cite{Hennessy:1988:ATP:50497}. Note that it is necessary to use the
failures-refinement relation in this condition, and not the trace-refinement
relation:~$(Q\parallel[\Sigma] U_F(p)) \hide \Sigma$ may have  the same
visible traces $\varepsilon$ and $\langle \epass\rangle$ as the ``Test Passed
Process'' $(\epass\then\Stop)$. However, the former may nondeterministically
refuse $\epass$, due to a deadlock occurring when a faulty SUT process
executes concurrently with $U_F(p,s)$ executing branch (\ref{eq:ufd}),
because $\#s = p$. This is explained further in the next paragraphs.

Intuitively speaking, $U_F(p)$ is able to perform any trace $s$ of $P$, up to
a length $p$. If, after having already run through $s\in\trc(P)$ with $\#s <
p$, an event is accepted by the SUT that is outside the initials of $P/s$,
the test immediately terminates with FAIL-event $\efail$. This is handled by
the branch (\ref{eq:ufa}) of the external choice in the process $U_F(p,s)$
defined above.

If $P/s$ is the $STOP$ process, this is revealed by its initials being empty.
In this case, the test may terminate successfully (branch (\ref{eq:ufb}) of
the external choice in $U_F(p,s)$). Note that at the same time, any (illegal)
event of the alphabet is also accepted by the test in branch~(\ref{eq:ufa}).
So, if the SUT accepts an event in a state where $P/s$ is supposed to have
stopped, there exists a test execution that terminates with FAIL by choosing
the first branch of the external choice.

If the length of $s$ is still less than $p$, the test accepts any event from
the initials $[P/s]^0$ and continues recursively as $U_F(p,s.e)$ in
branch~(\ref{eq:ufc}). A test of this type is called \emph{adaptive}, because
it accepts any legal behaviour of the SUT and adapts its consecutive
behaviour to the event selected by the SUT.

@@

\begin{theorem}\label{th:failurestest}
Let $P$ be a divergence-free CSP process over alphabet $\Sigma$
whose normalised transition graph $G(P)$ has $p$ states. Define fault domain ${\cal D}$ as
the set of all divergence-free CSP processes over alphabet $\Sigma$, whose transition graph
has at most $q$ states with $q \ge p$.
Then the test suite
\[
\TS_F = \{ U_F(k)~|~0 \le k < pq  \}
\]
is complete with respect to ${\cal F} = (P,\lessdet_F,{\cal D})$.
\end{theorem}
\begin{proof}
To prove soundness of $\TS_F$, suppose that $Q\in{\cal D} \wedge P\lessdet_F
Q$. In this case, $\trc(Q)\subseteq \trc(P)$, and Lemma~\ref{lemma:hseta}
implies that for all traces $s$ of $Q$, every $H$ in $\text{minHit}(P/s)$ is
a hitting set for $\minaccs(Q/s)$. So, when running in parallel with $Q$, any
adaptive test $U_F(p)$ will always enter the branches (\ref{eq:ufb}),
(\ref{eq:ufc}),  or (\ref{eq:ufd}) of the external choice construction for
$U_F(p,s)$. Branch (\ref{eq:ufb}) leads always to a PASS verdict, and branch
(\ref{eq:ufc}) to test continuation without a verdict. For the last branch,
we note that any selected minimal hitting set $H\in\text{minHit}(P/s)$ has a
non-empty intersection with each of the minimal acceptances of $Q/s$. As a
consequence, $Q/s$ never blocks when offered events from $H$, and the test
terminates with PASS event $\epass$. Note that this argument requires that
$Q$ is free of livelocks, because otherwise the PASS-events might not become
visible, due to unbounded sequences of hidden events performed by $Q$. Note
further, that this proof did not refer in any way to the size of $Q$'s
normalised transition graph, so the test suite is sound for {\it all}
non-divergent CSP process refining $P$. \fixme{alcc: Soundness does not
depend on the fault domain. I would have it as a separate theorem.}

To prove exhaustiveness, consider a process $Q\in{\cal D}$ with
$P\not\lessdet_F Q$. This non-conformance can be caused in two possible ways.
\begin{description}
\item[Case~1] $\trc(Q)\not\subseteq \trc(P)$
\item[Case~2] There exists a joint trace $s\in\trc(Q)\cap\trc(P)$ and a minimal acceptance $A_Q$
of $\minaccs(Q/s)$, such that
(see Lemma~\ref{lemma:tgtrcref}, (\ref{eq:failrefb})).
\begin{equation}
\label{eq:accsnotcontained}
\forall A_P\in\minaccs(P/s): A_P\not\subseteq A_Q,
\end{equation}
\end{description}
It has to be shown for each of the two possibilities that at least one test
execution of some $(Q\parallel[\Sigma] U_F(k))$ with $k < pq$ ends with the
FAIL event $\efail$ or without giving any verdict. The latter case is also
interpreted  as FAIL, since then the process $\epass\then\Stop$ is no longer
failures-refined by the test execution.

For the first case, consider a  trace $s.e\in\trc(Q)$ such that
$s\in\trc(P)$, but $s.e\not\in\trc(P)$. Such a trace always exists because
$\varepsilon$ is a trace of every process. In this case, $s$ is also a trace
of the product graph $G = G(P)\times G(Q)$ defined in Section~\ref{sec:ntg}.
From the construction of $G$ described there, we know that $G$ has at most
$pq$ reachable states, because $G(P)$ has $p$ states, and $G(Q)$ has at most
$q$ states. Suppose that $G/s = (n_P,n_Q)$. Applying
Lemma~\ref{lemma:extendV} implies that this state can be reached by a trace
$u\in\trc(G)$ of length $\#u < pq$. Now the construction of the transition
function of $G$ implies that $u$ is also a trace of $P$ and $Q$. Since test
$U_F(pq-1)$ accepts all traces of $P$ up to length $pq-1$, $u$ is also a
trace of this test, and, by construction, $U_F(pq-1)/u = U_F(pq-1,u)$. Since
$s.e\not\in\trc(P)$, $e$ is an element of $\Sigma-[P/u]^0$. Therefore, in at
least one execution, $U_F(pq-1,u)$ executes its first branch (\ref{eq:ufa})
with this event $e$, so that the test fails with event $\efail$. Again, the
assumption of non-divergence of Q is needed for this conclusion. \fixme{alcc:
I didn't see how Lemma 3 is being applied here. Also, given the definition of
the test as a process, I'm not sure why you need to make the argument using
graph product.}

For the Case~2, we note that trace $s$ is again a trace of the product graph
$G$. Therefore, $G/s$ can again be reached by a trace
$u\in\trc(Q)\cap\trc(P)$ of maximal length $\#u < pq$. Consider test $U_F(\#
u)$, which satisfies $U_F(\# u)/u = U_F(\#u,u)$, because it always performs
branch (\ref{eq:ufc}) until the trace $u$ has been completely processed.
$U_F(\#u,u)$ may execute branches (\ref{eq:ufa}) or (\ref{eq:ufd})
only:~assumption (\ref{eq:accsnotcontained}) in Case~2 implies that $P/s$ has
at least one non-empty minimal acceptance, so the guard condition $([P/s]^0 =
\varnothing)$ of branch (\ref{eq:ufb}) evaluates to $\isf$ for $U_F(\#u,u)$.
Moreover, the guard condition $(\#s < p)$ for branch (\ref{eq:ufc}) evaluates
to $\isf$ for $U_F(\#u,u)$, too. If branch (\ref{eq:ufa}) is executed, the
test always fails. If branch (\ref{eq:ufd}) is executed, the test fails for
the execution where a minimal hitting set $H\in\text{minHit}(P/u)$ is chosen
by $U_F(\#u,u)$ that has an empty intersection with the minimal acceptance
$A_Q$ from condition (\ref{eq:accsnotcontained}). The existence of such an
$H$ is guaranteed because of Lemma~\ref{lemma:hseta}. As a consequence, there
exists a test execution   where $Q/u$ selects acceptance $A_Q$ and
$U_F(\#u,u)$ selects $H$. This execution deadlocks in process state
$(Q\parallel[\Sigma]U_F(\# u))/u$, so it cannot produce the pass event
$\epass$; this  means that the test fails. This concludes the proof. \xbox
\end{proof}
